\documentclass{beamer}

\usepackage[latin1]{inputenc}
\usepackage{cmll}
\usepackage{lstmacros}
\usepackage{graphicx}
\usepackage{array}
\usepackage{subfigure}
\usepackage{amsmath}
\usepackage{comment}
\usepackage{etoolbox}
\usepackage[cal=cm,calscaled=.80]{mathalfa}
\usepackage{tikz}
\usetikzlibrary{arrows,positioning}
\usepackage[style=alphabetic,natbib=true]{biblatex}

\usetheme{Warsaw}

\newtoggle{tdlp}
\toggletrue{tdlp}
\input{propmacros}
\newcommand{\altt}{~|~}

\let\oldframetitle\frametitle
\renewcommand{\frametitle}[1]{
  \oldframetitle{#1}\vspace{-3.5cm}
}
\newcommand{\ftitle}[2]{\oldframetitle{#1}\vspace{-#2cm}}

\setbeameroption{show notes}
\setbeamertemplate{note page}[plain]

\newcommand\CPFSize{\fontsize{7}{8.4}\selectfont}
% CP/GV example listing details
\lstdefinelanguage{CP}{
    mathescape=true,
    morekeywords=[1]{case, inl, inr},
    sensitive=false,
    morecomment=[l]{//},
    morecomment=[s]{/*}{*/},
    morestring=[b]",
    basicstyle=\CPFSize,
    keywordstyle=[1]{\color{red}},
}
\lstdefinelanguage{GV}{
    mathescape=true,
    morekeywords=[1]{case, connect, in, let, of, receive, select, send,
      terminate, to, with},
    sensitive=false,
    morecomment=[l]{//},
    morecomment=[s]{/*}{*/},
    morestring=[b]",
    basicstyle=\HappyMed,
    keywordstyle=[1]{\color{magenta}},
}

\lstnewenvironment{cp}{\lstset{language=CP}}{}
\lstnewenvironment{gv}{\lstset{language=GV}}{}

\title[Project Presentation]{Propositions as Sessions, Mechanically}
\author{Craig McLaughlin}
\institute{University Of Glasgow}
\date{\today}

\begin{document}

\begin{frame}
\titlepage
\note{
today i am going to talk about a formalisation effort based on an existing
type system whose properties have previously only been studied
informally. first in order to motivate the previous and current work i shall
make two general observations here. one: distributed communication systems are
becoming increasingly prevalent in modern society. two: the correct
functioning of these systems hinges on the protocol between any two
communicating agents to be well-defined and correctly implemented.
}

\end{frame}

\section{Preamble}

\begin{frame}
\ftitle{Session Types}{1.5}
\begin{math}
  \begin{array}{cccc}
    S & \bnf  & \inpt{T}{S} & \mbox{receive value of type T, continue as S}\\
      & \altt & \outpt{T}{S} & \mbox{send value of type T, continue as S}\\
      & \altt & \Branch{\mathcal{L}_i}{S_i}{i\in I} &
                           \mbox{offer between $|I|$ alternatives}\\
      & \altt & \Choice{\mathcal{L}_i}{S_i}{i\in I} &
                           \mbox{choose between $|I|$ alternatives}\\
      & \altt & \mbox{\lstinline{end}} & \mbox{terminate session}\\
 \dual{S} & \bnf & \mbox{dual of S}\\
 \dual{(\inpt{T}{S})} & \bnf & \outpt{T}{\dual{S}}\\
 \dual{\Branch{\mathcal{L}_i}{S_i}{i\in I}} & \bnf &
       \Choice{\mathcal{L}_i}{\dual{S_i}}{i\in I}\\
            \vdots                & \cdots & \vdots
 \end{array}
\end{math}
\note{
one approach to achieve such correctness guarantees is by using binary session
types. [INTRODUCE BINARY SESSION TYPES HERE]. specifying a communication
protocol then involves providing mutual dual types to the communicating
parties.
}
\end{frame}

\begin{frame}
\ftitle{System Overview}{2}
CP $=$ process calculus with operators from classical linear logic

GV $=$ functional language with session types

GV $\to$ CP translation
\\~\\
Relies on Curry-Howard variant:
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}ccc}
Propositions & correspond to & Session Types\\
Proofs & correspond to & Processes\\
Cut Elimination & correponds to & Communication
\end{tabular*}
\note{
the formalisation i present here is an encoding of a system in the Coq proof
assistant. it is based on a two-tier system providing a logical foundation for
understanding session types. A high-level overview of the system is as
follows: we have a process calculus CP with operators from classical linear
logic, a high-level functional language GV with session types, and a
translation from GV to CP which interprets operators in classical linear logic
as session types.
}
\end{frame}

\begin{frame}[fragile]
\ftitle{CP Example}{0.5}
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} cc}
Process A ($A_x$) & Process B ($B_x$)\\
$\colored\tp{Int \parr (String \otimes \1)}$
  ($\colored\tp{\dualbt{S}}$) &
$\colored\tp{Int \otimes (String \parr \bot)}$ ($\colored\tp{S}$)\\
\begin{cp}
$\colored\tm{x(y).x[w].(\mbox{put-thanks}_w \mid x[\,].0)}$
\end{cp} &
\begin{cp}
$\colored\tm{x[y].(\mbox{put-gift}_y \mid x(w).x().\mbox{get-response}_w)}$
\end{cp}
\end{tabular*}
\[\colored
\inference{
  \tm{B_x} \tpvdash \tp{\Gamma \comma \tmof{x}S}
  &
  \tm{A_x} \tpvdash \tp{\Delta \comma \tmof{x}\dualbt{S}}
}{Cut}{
  \tm{\nu x \of{S}.(B_x \mid A_x)} \tpvdash \tp{\Gamma \comma \Delta}
}
\]
\CPFSize
\colored
\begin{align*}
  \tm{\nu x.(B_x \mid A_x)} & =
    \tm{\nu x.(x[y].(\mbox{put-gift}_y \mid x(w).x().\mbox{get-response}_w)
          \mid x(y).x[w].(\mbox{put-thanks}_w \mid x[\,].0))}
    \\
                            & \becomes
    \tm{\nu y.(\mbox{put-gift}_y \mid
               \nu x.(x(w).x().\mbox{get-response}_w) \mid
                      x[w].(\mbox{put-thanks}_w \mid x[\,].0))}
    \\
                            & \becomes
    \tm{\nu y.(\mbox{put-gift}_y \mid
               \nu w.(\mbox{put-thanks}_w \mid
                      \nu x.(x().\mbox{get-response}_w \mid x[\,].0)))}
    \\
                            & \becomes
    \tm{\nu y.(\mbox{put-gift}_y \mid
               \nu w.(\mbox{put-thanks}_w \mid \mbox{get-response}_w))}
\end{align*}
\note{
Here are some exmaples of linear logic connectives with session type
interpretations. The operator times corresponds to output and the operator par
corresponds to input. Since it is based on the classical variant of linear
logic we have that negation is an involution.
}
\end{frame}

\begin{frame}[fragile]
\ftitle{GV Example}{2}
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} cc}
Process A ($A_x$) & Process B ($B_x$)\\
$\colored\tp{\inpt{Int}{\outpt{String}{\key{end}_!}}}$ ($\colored\tp{S}$) &
$\colored\tp{\outpt{Int}{\inpt{String}{\key{end}_?}}}$
  ($\colored\tp{\overline{S}}$)\\
\begin{gv}
let (m,$x_1$) = receive $x$ in
send thanks $x_1$
\end{gv} &
\begin{gv}
let $x_1$ = send gift $x$ in
let (resp,$x_2$) = receive $x_1$ in
terminate $x_2$
\end{gv}
\end{tabular*}
\[\colored
\inference{
  \Tp{thanks:\tp{String} \comma x:\tp{S}} \Tpvdash A_x:\tp{\key{end}_!} &
  \Tp{gift:\tp{Int} \comma x:\tp{\overline{S}}} \Tpvdash B_x:\tp{\key{Unit}}
}{Connect}{
  \Tp{thanks:\tp{String} \comma gift:\tp{Int}} \Tpvdash
    \Tm{with}~x~\Tm{connect}~A_x~\Tm{to}~B_x : \tp{Unit}
}
\]
\end{frame}

\begin{frame}
\frametitle{Motivation}
Informal proofs $=$ Error-prone $+$ Difficult to update
\begin{itemize}
\item Subject Reduction: If $m : T$ and $m \rightarrow m'$ then $m' : T$
\end{itemize}
\note{
there has been a lot of activity in this area producing a variety of different
session-based type systems. many systems build on top of previous work adding
more complex features such as asynchronous, polymorphic and recursive session
types. extending informal proofs is error-prone and difficult because it is
not always clear which parts of a system's metatheory is affected by a
change. a formalised system would reduce likelihood of errors in the proofs
and aid extension since the theorem prover would highlight changes to the
metatheory
}
\end{frame}

\section{Formalisation}

\begin{frame}
\frametitle{Goals}
\begin{itemize}
\item Reuse where possible
\item Minimise difference between informal/formal variants
\item Avoid known pitfalls
\end{itemize}
\note{
The encoding follows practices in formalising programming language metatheory
with a principal design goal of reusing existing libraries from previous
developments. Additionally, I want to keep as close a correspondence as
possible between the development and the pen-and-paper presentation in order
to mitigate questions of adequacy, you know, whether the formal system is what
you intended to formalise. And lastly, I aimed to exploit the features of the
Coq system without over complicating the development such as taking careful
consideration over representation choices.
}
\end{frame}

\begin{frame}
\oldframetitle{Preliminaries}
\begin{itemize}
\item Locally Nameless
  \begin{itemize}
  \item Avoids variable capture
  \item de Bruijn indices for bound variables
    ($\lambda~(\lambda~1) \equiv \lambda x. \lambda y. x$)
  \end{itemize}
\item Cofinite Quantification
  \begin{itemize}
  \item stronger induction principle
  \item
    \begin{math}
    \begin{array}{lcl}
      \mbox{Exists-Fresh} & & \mbox{Cofinite}\\
      \inference{
        x \notin FV(N) &
        \Tp{\Gamma \comma \Tmof{x}T} \Tpvdash \Tmof{N^x}\Tp{U}
      }{}{
        \Tp{\Gamma} \Tpvdash \Tmof{abs~N}\Tp{T \to U}
      }
    & \mbox{vs}. &
      \inference{
        \forall x \notin L.
        (\Tp{\Gamma \comma \Tmof{x}T} \Tpvdash \Tmof{N^x}\Tp{U})
      }{}{
        \Tp{\Gamma} \Tpvdash \Tmof{abs~N}\Tp{T \to U}
      }
    \end{array}
    \end{math}
  \end{itemize}
\item Metatheory library for representation
\end{itemize}
\note{
one of the key issues when formalising programming language metatheory is how
to represent variables both bound and free. I chose the locally nameless
representation which separates bound and free variables into distinct
entities. Thus it avoids the variable capture problem in contrast to other
approaches.

In this approach, a de Bruijn index is a number used to represent a bound
variable which indicates the position of its binder (starting from zero for
the innermost binder).

Constructs with binders have to open the term with a free variable and this
requires some technique for handling freshness of the opening variable. I
chose to adopt the cofinite quantification approach which allows to exclude a
finite set of names from consideration in contrast to the sometimes weaker
notion of a single sufficiently fresh name

I utilise the metatheory library developed by a team of researchers at the
university of pennslevennia which handles these representation choices.
}
\end{frame}

\subsection{CP}

\begin{frame}[fragile]
\oldframetitle{CP Rules: Formal vs. Informal Comparison}
\begin{tabular}{lc}
\begin{coqp}
Inductive cp_rule : proc -> penv -> Prop :=
  | $\cdots$
  | cp_cut :
      forall (L:atoms) P Q A $\Gamma$ $\Delta$P $\Delta$Q
             (PER: Permutation $\Gamma$ ($\Delta$P ++ $\Delta$Q))
             (CPP: forall (x:atom) (NL: x $\notin$ L),
                     (open_proc P x) $\tpcp$ (x $\sim$ A) ++ $\Delta$P)
             (CPQ: forall (x:atom) (NL: x $\notin$ L),
                (open_proc Q x) $\tpcp$ (x $\sim$ $\dualbt{A}$) ++ $\Delta$Q),
        $\nu$ A.(P $\mid$ Q) $\tpcp$ $\Gamma$
  | $\cdots$
\end{coqp} &
\begin{math}
\colored
\inference{
  \tm{P} \tpvdash \tp{\Gamma \comma \tmof{x}A}
  &
  \tm{Q} \tpvdash \tp{\Delta \comma \tmof{x}\dualbt{A}}
}{}{
  \tm{\nu x \of{A}.(P \mid Q)} \tpvdash \tp{\Gamma \comma \Delta}
}
\end{math}
\end{tabular}
\note{

CP rules were encoded as an inductive in Coq and an example for cut is shown
here. CP provided challenges for the metatheory library since it differs in
many respects to lambda calculi. for example, in the lambda calculus one does
not work up to permutation of binders whereas in a process calculus these
kinds of structural equivalences are standard as we shall see. we require
additional assumptions for environment ordering and more attention is paid to
the opening of subprocesses. the metatheory library did not support
permutations and as a result of this work some basic operations and custom
tactics were developed. the custom tactics can solve simple permutation goals
via simple proof search. this slide presents the informal and formal
definitions of cut. notice there is no overhead when introducing binders to
the subprocesses or for environment ordering, but apart from these
representation issues the rules look similar which could assist the uptake of
such formal tools by programming language researchers.

}
\end{frame}

\begin{frame}[fragile]
\oldframetitle{Subject Reduction in CP}
\begin{tabular}{ll}
\begin{coq}
Theorem proc_sub_red:
  forall $\Gamma$ P Q
         (WT: P $\tpvdash_{cp}$ $\Gamma$)
         (RED: P $\becomes$ Q),
    Q $\tpvdash_{cp}$ $\Gamma$.
\end{coq} &
\begin{math}
\becomes \defeq \left\{ \begin{array}{ll}
                   \mbox{Principal cut reductions}~(\beta)\\
                   \mbox{Commuting Conversions}~(\kappa)
                   \end{array} \right.
\end{math}
\end{tabular}
\[\colored
(\kappa_{\otimes1}) \quad
\tm{\nu z\Of{C}.(x[y].(P \mid Q) \mid R)}
\quad \becomes \quad
\tm{x[y].(\nu z\Of{C}.(P \mid R) \mid Q)}
\]
\note{
Subject reduction in CP proceeds by cut reductions of the logical operators
against their duals and \textit{commuting conversions} as well as some other
rules. Commuting conversions push a cut inside a logical operator and the
intuition is that they permit further communication to occur which was guarded
by a cut.
}
\end{frame}

\begin{frame}[fragile]
\oldframetitle{Principal Cut Reduction ($\beta_{!?}$): Formal vs. Informal}
\begin{tabular}{c}
\begin{coq}
Lemma reduce_spawn:
  forall P Q A dA $\Gamma$
         (DUA: dual_props A dA)
         (WT: $\nu$ !A.(!0(A).P $\mid$ ?0[dA].Q) $\tpcp$ $\Gamma$),
    $\nu$ A.(P $\mid$ Q) $\tpcp$ $\Gamma$.
\end{coq}
\\~\\
\begin{math}
\begin{array}{c}
\small
\colored
\inference{
  \inference{
    \tm{P} \tpvdash \tp{{?\Gamma} \comma \tmof{y}A}
  }{!}{
    \tm{!x(y).P} \tpvdash \tp{{?\Gamma} \comma \tmof{x}{!A}}
  }
  &
  \inference{
    \tm{Q} \tpvdash \tp{\Delta \comma \tmof{y}\dualbt{A}}
  }{?}{
    \tm{?x[y].Q} \tpvdash \tp{\Delta \comma \tmof{x}{?\dualbt{A}}}
  }
}{}{
  \tm{\nu x \Of{!A}.(!x(y).P \mid {?x[y].Q})} \tpvdash
    \tp{{?\Gamma} \comma \Delta}
}
\becomes
\inference{
  \tm{P} \tpvdash \tp{{?\Gamma} \comma \tmof{y}A}
  \\
  \tm{Q} \tpvdash \tp{\Delta \comma \tmof{y}\dualbt{A}}
}{}{
  \tm{\nu y\Of{A}.(P \mid Q)} \tpvdash \tp{{?\Gamma} \comma \Delta}
}
\normalsize
\end{array}
\end{math}
\end{tabular}
\note{
An example of a principal cut reduction: cut of server accept against client
request. Comparing the formal and informal versions, the informal version
describes the cut reductions using derivation trees while this structure is
left implicit in the Coq encoding.
}
\end{frame}

\begin{frame}[fragile]
\oldframetitle{Commuting Conversion ($\kappa_{\otimes1}$): Formal vs.
  Informal}
\begin{tabular}{c}
\begin{coq}
Lemma reduce_cc_multi_one:
  forall P Q R (x:atom) A B $\Gamma$ (LCQ: lc_proc Q)
         (WT: $\nu$ A.(x[B].(P $\mid$ Q) $\mid$ R) $\tpcp$ $\Gamma$),
    x[B].($\nu$ A.($\permb{0}{1}{P}$ $\mid$ R) $\mid$ Q) $\tpcp$ $\Gamma$.
\end{coq}
\\~\\
\begin{math}
\colored
\begin{array}{c}

% \spacer
\vcenter{
  \inference{
    \inference{
      \tm{P} \tpvdash \tp{\Gamma \comma \tmof{y}A \comma \tmof{z}C}
      &
      \tm{Q} \tpvdash \tp{\Delta \comma \tmof{x}B}
    }{\otimes}{
      \tm{x[y].(P \mid Q)} \tpvdash
        \tp{\Gamma \comma \Delta \comma \tmof{x}A \otimes B \comma \tmof{z}C}
    }
    &
    \tm{R} \tpvdash \tp{\Theta \comma \tmof{z}C^\bot}
  }{Cut}{
    \tm{\nu z\Of{C}.(x[y].(P \mid Q) \mid R)} \tpvdash
      \tp{\Gamma \comma \Delta \comma \Theta \comma \tmof{x}A \otimes B}
  }
}
\quad \becomes \hfill \\~\\
\hfill
\vcenter{
  \inference{
    \inference{
      \tm{P} \tpvdash \tp{\Gamma \comma \tmof{y}A \comma \tmof{z}C}
      &
      \tm{R} \tpvdash \tp{\Theta \comma \tmof{z}C^\bot}
    }{Cut}{
      \tm{\nu z\Of{C}.(P \mid R)} \tpvdash
        \tp{\Gamma \comma \Theta \comma \tmof{y}A}
    }
    &
    \tm{Q} \tpvdash \tp{\Delta \comma \tmof{x}B}
  }{\otimes}{
    \tm{x[y].(\nu z\Of{C}.(P \mid R) \mid Q)} \tpvdash
      \tp{\Gamma \comma \Delta \comma \Theta \comma \tmof{x}A \otimes B}
  }
}

\end{array}
%% \colored
%% (\kappa_{\otimes1}) \quad
%% \tm{\nu z\Of{C}.(x[y].(P \mid Q) \mid R)}
%% \quad \becomes \quad
%% \tm{x[y].(\nu z\Of{C}.(P \mid R) \mid Q)}
\end{math}
\end{tabular}
\note{

(well, that looks ridiculous.)\\~\\

Comparing formal and informal versions of a commuting conversion for output
reveals some issues. it is necessary to stipulate that Q is locally closed to
ensure that the P subprocess receives the channel from the outer cut as part
of its environment -- since the environment splitting is in general
non-deterministic. Additionally, we need to permute the binders of P in the
conclusion and this is necessary for the majority of the commuting conversion
rules.

}
\end{frame}

\begin{frame}[fragile]
\oldframetitle{Top-Level Cut Elimination}
\begin{coq}
Theorem proc_cut_elim:
  forall P $\Gamma$
         (WT: P $\tpcp$ $\Gamma$),
    exists Q, P $\becomes^\star$ Q /\ $\sim$ is_cut Q.
\end{coq}
\note{
mention star being the reflexive and transitive closure of reduction

Well-typed relation alone is insufficient to prove the theorem. The rule for
cut does not specify the subprocesses are related i.e. are eventually dual
logical operators which would be necessary to proceed using the current
definition.

It seems necessary to define a termination measure on the length of the
derivation but this has not yet been done. Additionally, work on formalising
logics with cut elimination take an entirely different approach favouring an
explicit representation of derivations as trees. Adopting this approach for CP
may prove fruitful not just in this instance but also in other areas of the
development such as specifying environment splitting as in the rules for
commuting conversions.

}
\end{frame}

\subsection{GV}

\begin{frame}
GV is similar
\note{

Typing rules again? Very similar to CP.\\~\\
Representation issues?\\~\\
GV has built-in support for session types. consider the rule for sending a
value of type T along a channel N of type !T.S.

}
\end{frame}

\subsection{Translation}

\begin{frame}
something about a translation

\note{

the translation to CP is
presented. blah blah something about the session types being a little odd or
overlook this probably

the translation seems to require a duplication of the specifications for cp
and gv typing rules not done

}
\end{frame}

\subsection{Issues}

\begin{frame}
\frametitle{Issues and Guidelines}
\begin{itemize}
\item representation
\item modularisation
\item automation
\item reuse
\end{itemize}
\note{
  representation, modularisation(?), automation, reuse
}
\end{frame}

\section{Future Work}

\begin{frame}
\oldframetitle{Future Work}
\begin{itemize}
\item difficulties in formalisation
\item implications
\item future work
\end{itemize}
\note{


}
\end{frame}

\printbibliography

\end{document}
